### 2023-12-15 sections/1-introduction.tex
edits to introduction

Often dubbed ``the Big Data explosion,'' rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$.

Biology has also seen an influx of large, high-dimensional datasets.
The GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences, each 7,682 characters in length;
Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 million genomes with aligned length of 50,000 letters.

@@

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating that computer architecture will not ``catch up'' to our computational needs in the near future.
As a result, analysis of these large datasets requires better algorithms.

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4~\cite{OpenAI2023GPT4TR}, and LLAMA-2~\cite{Touvron2023Llama2O}, similarity search can become a useful tool for understanding and verifying the validity of model embeddings. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Previous works have used the term \emph{approximate} search to refer to $\rho$-nearest neighbor search, but in this paper, 
we reserve the term \emph{approximate} for $k$-NN search algorithms which may miss some of the $k$ nearest elements, and erroneously include some of the elements other than the truly nearest $k$.
In other words, we call a $k$-NN search algorithm \emph{approximate} if there exists a query, dataset, distance function, or value of $k$ for which its set $H$ of returned points satisfies the condition $H \triangle S_q \neq \emptyset$, where $\triangle$ denotes a symmetric difference.
When we refer to an \emph{exact} $k$-NN algorithm, we describe an algorithm which, no matter the query, dataset, distance function, nor value of $k$, returns precisely the $k$ nearest neighbors to that query;
no closer points are elided and no further points are included.

$k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Na\"{i}ve implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not exploit the geometric and topological structure inherent in large datasets.
Approximate search is sufficient for some applications, but the need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$, and small values of $k$ may be necessary when classes are not well-separated.
There is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

@@

This paper focuses on four novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-nearest neighbors search algorithms in CHESS, as well as improved genericity across distance functions.
Additionally, we prove a guarantee about the scaling behavior of cluster radii. 
Finally, we provide a comparison of CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} and HNSW, on several datasets from the ANN benchmarks suite~\cite{aumuller2020ann}.

>>>>

Often dubbed ``the Big Data explosion,'' rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
% Najib: We should mention datasets in this order: neural -network embeddings, biological, radio-ml, and then astronomical. DONE. (flow is a bit awakward though)

With the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4~\cite{OpenAI2023GPT4TR}, and LLAMA-2~\cite{Touvron2023Llama2O}, neural network embeddings are becoming an important source of big data. 
Biology has also seen an influx of large, high-dimensional datasets.
The GreenGenes project~\cite{desantis2006greengenes} provides a multiple-sequence alignment of over one million bacterial 16S sequences, each 7,682 characters in length;
Silva 18S~\cite{10.1093/nar/gks1219} contains ribosomal DNA sequences of approximately 2.25 million genomes with aligned length of 50,000 letters.

Radio frequency data offers another example of massive datasets. 
For example, the RadioML~\cite{oshea2018radioml} dataset contains 2,555,504 samples of both live and synthetically generated signal captures of different modulation modes over a range of SNR levels.

Additionally, many large, high-dimensional datasets come from the field of astronomy. The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions.

@@

In many fields, the sizes of datasets are growing exponentially, and this increase in the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}.
This indicates that computer architecture will not ``catch up'' to computational needs in the near future.
Thus, analysis of these large datasets requires better algorithms.

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation~\cite{annoy} and classification systems~\cite{suyanto2022knnclassifier}.

As the cardinalities and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search 
aims to find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Previous works have used the term \emph{approximate} search to refer to $\rho$-nearest neighbor search, but in this paper, 
we reserve the term \emph{approximate} for search algorithms which do not exhibit perfect recall.
When we refer to an \emph{exact} search algorithm, we mean an algorithm which exhibits perfect recall when compared to a na\"{i}ve linear search.

$k$-nearest neighbors ($k$-NN) search is one of the most ubiquitous classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Na\"{i}ve implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets because their cardinalities are growing exponentially.

While algorithms for fast $k$-NN on large datasets do exist, most do not exploit the geometric and topological structure inherent in large datasets.
Further, such algorithms are often approximate and while approximate search may be sufficient for some applications, the need for efficient and \emph{exact} search remains.
For example, for a majority voting classifier, approximate $k$-NN may agree with exact $k$-NN for large values of $k$, but may be sensitive to local perturbations with smaller values of $k$.
This is especially true when classes are not well-separated~\cite{zhang2022imbalanced}.
Further, there is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

@@

This paper focuses on four novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-nearest neighbors search algorithms in CHESS, as well as improved genericity across distance functions.
Finally, we provide a comparison of CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion}, HNSW~\cite{malkov2016hnsw}, and ANNOY~\cite{annoy} on several datasets from the ANN benchmarks suite~\cite{aumuller2020ann}.

### 2023-12-15 sections/3-datasets.tex
headers for new datasets and naming fixes
edits to introduction

@@

>>>>

\subsection{ANN Benchmark Datasets}
\label{subsec:ann-benchmarks}

@@

\subsection{Silva}
\label{subsec:silva}

Add section about Silva.

\subsection{Radio ML}
The RadioML dataset contains samples of both live and synthetically generated signal captures of different modulation modes over a range of SNR levels. Specifically, it is comprised of 24 modulation modes at 26 different SNR levels ranging from -20 dB to 30 dB, with 4,096 samples at each modulation mode and SNR level~\cite{oshea2018radioml}. Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total. Each sample provided in RadioML can be thought of as an array of 2048 floating point values for a single signal capture with labels for SNR and modulation mode.

\label{subsec:radioml}
